Single-seismometer Focal Mechanism and Uncertainty Estimation from Body Waves

### 2024-10-29 paper/main.tex
latex updates
Done with tolerance, may revisit during uncertainty

% import packages
\usepackage{gensymb, nicefrac}

@@

    [Suzan] Insert words here...\\
    Talk about this being an inverse method for a highly nonlinear problem\\
    Writing tip: maximize information content of words

@@

    \subsection{Derivation}
     When a quake event happens, a single seismometer reads $P$, $SV$ and $SH$ body waves through
     which the respective amplitudes $A^P$, $A^{SV}$ and $A^{SH}$ along with their noise levels
     $\sigma^P$, $\sigma^{SV}$ and $\sigma^{SH}$ (representing standard deviation of an assumed normal
     distribution centered at the amplitude value) can be extracted.
     Given a ray path's azimuth $\phi$ and take-off angles $i$, $j$ of $P$ and $S$ waves respectively,
     we model the amplitudes as functions of the source mechanism's fault strike $\psi$, fault dip
     $\delta$ and slip rake $\lambda$, further accounting for the $P$- and $S$-velocities $\alpha_h$,
     $\beta_h$ at source depth $h$ \citep{AkiRichards1980}. The expressions are:

@@

     We thus need a metric against which to evaluate what the best and acceptable estimates are. The
     misfit function used by \citet{sita_potential_2022} is defined as the angle between the vector of
     observed amplitudes and that synthesized from parameter estimates. In this paper, we shall use a
     misfit $\mu$ of cosine similarity between the two vectors which has a 1-to-1 correspondence with angle:

     \begin{align} \label{eq:eq3}
         \mu(\mathbf{p}) = \frac{\mathbf{A}_s^\top\mathbf{A}_o}{\|\mathbf{A}_s\|\|\mathbf{A}_o\|}
     \end{align}

     where $\mathbf{A}_o = (A^P, A^{SV}, A^{SH})^\top$ are observed amplitudes, $\mathbf{A}_s =
     (\widehat{A}^P, \widehat{A}^{SV}, \widehat{A}^{SH})^\top$ are synthetic amplitudes, and
     $\mathbf{p} = (\widehat{\psi}, \widehat{\delta}, \widehat{\lambda})$ are parameter estimates.

     \subsection{Justification}
     We are interested in situations where absolute amplitudes are either unknown or unnecessary, for
     instance if using data collected from NASA's Insight mission \citep{sita_potential_2022}.
     In this case, we cannot know the absolute velocities because they depend on an understanding of
     Mars' interior that is good enough to build a reliable velocity model. Alternative misfit
     functions can account for the relative amplitudes, such as in Equations \ref{eq:eq4} and
     \ref{eq:eq5}:

     \begin{align} \label{eq:eq4}
         \mu(\mathbf{p}) = \sqrt{\left( \Delta \frac{A^{P}}{A^{SV}}\right)^2 + \left( \Delta \frac{A^{SV}}{A^{SH}}\right)^2}
     \end{align}

     \begin{align} \label{eq:eq5}
         \mu(\mathbf{p}) = \sqrt{\log^2 \left( \Delta \frac{A^P}{A^{SV}} \right) + \log^2 \left( \Delta \frac{A^{SV}}{A^{SH}} \right)}
     \end{align}

     First, we note that using cosine similarity is more numerically stable in cases where one of the
     observed or synthetic amplitudes is close to zero since the norm of the associated vector in
     Equation \ref{eq:eq3} may still be large enough to make division feasible.

     Secondly, we have chosen cosine similarity over angle because it involves one less operation,
     which is more computationally efficient and easier to differentiate for directed search
     algorithms (see Section \ref{sec:algorithms}). The Gaussian function in Equation \ref{eq:eq6} is
     similarly easy to differentiate, but it does not account for relative amplitudes so it would
     exclude otherwise acceptable solutions.

     \begin{align} \label{eq:eq6}
         &\mu(\mathbf{p}) \sim \exp \left( - \frac{1}{2} (\mathbf{A}_s - \mathbf{A}_o)^\top \mathbf{\Sigma}^{-1} (\mathbf{A}_s - \mathbf{A}_o) \right)\\
         \nonumber &\text{where } \mathbf{\Sigma} = \text{diag}(\sigma^P, \sigma^{SV}, \sigma^{SH})
     \end{align}

     \subsection{Tolerance}
     Given vectors $\mathbf{A}_s$ and $\mathbf{A}_o$, we would like to know the largest cosine
     similarity for which $\mathbf{A}_s$ is considered an acceptable fit for the recorded amplitudes
     while respecting the asymmetry of error levels for different amplitudes. This is visualized in 3D
     space as a confidence ellipsoid centered at $\mathbf{A}_o$ (Equation \ref{eq:eq7}), for which
     acceptable fits are vectors that fall inside the elliptical cone with vertex at the origin and
     tangent to the ellipsoid.
     \begin{align} \label{eq:eq7}
         \sum_{k \in \{P, SV, SH\}} \frac{(A^k_b-A^k)^2}{(\sigma^k)^2} = 1
     \end{align}

     where $\mathbf{A} = (A^P_b, A^{SV}_b, A^{SH}_b)^\top$ describes a boundary point. The elliptical
     cone intersects with the ellipsoid at every boundary point for which
     \begin{align} \label{eq:eq8}
         \sum_{k \in \{P, SV, SH\}} \frac{2A^k_b(A^k_b-A^k)}{(\sigma^k)^2} = 0
     \end{align}

     Combining equations \ref{eq:eq7} and \ref{eq:eq8} gives the plane containing every point of 
     tangency:
     \begin{align} \label{eq:eq9}
         \sum_{k \in \{P, SV, SH\}} \frac{(A^k_b-A^k)^2}{(\sigma^k)^2} - \frac{2A^k_b(A^k_b-A^k)}{(\sigma^k)^2} = 1\\
         \nonumber \sum_{k \in \{P, SV, SH\}} \frac{A^k}{(\sigma^k)^2} A^k_b = \left( \sum_{k \in \{P, SV, SH\}} \frac{(A^k)^2}{(\sigma^k)^2} \right) -1
     \end{align}

     Let $\mathbf{N}_1 = (\nicefrac{A^P}{(\sigma^P)^2}, \nicefrac{A^{SV}}{(\sigma^{SV})^2},
     \nicefrac{A^{SH}}{(\sigma^{SH})^2})^\top$ be the normal of this plane. Consider also the plane
     containing vectors $\mathbf{A}_o$ and $\mathbf{A}_s$, with normal
     \begin{align} \label{eq:eq10}
         \mathbf{N}_2 = \mathbf{A}_o \times \mathbf{A}_s = (,,)^\top
     \end{align}

@@

>>>>

% import packages
\usepackage{gensymb, amssymb, nicefrac, bm}

@@

    [Suzan] Insert words here...\\
    Talk about inverse method for a highly nonlinear problem\\
    Writing tip: maximize information content of words

@@

    \subsection{Derivation}
    When a quake event happens, a single seismometer reads $P$, $SV$ and $SH$ body waves through
    which the respective amplitudes $A^P$, $A^{SV}$ and $A^{SH}$ along with their noise levels
    $\sigma_P$, $\sigma_{SV}$ and $\sigma_{SH}$ (representing standard deviation of an assumed normal
    distribution centered at the amplitude value) can be extracted.
    Given a ray path's azimuth $\phi$ and take-off angles $i$, $j$ of $P$ and $S$ waves respectively,
    we model the amplitudes as functions of the source mechanism's fault strike $\psi$, fault dip
    $\delta$ and slip rake $\lambda$, further accounting for the $P$- and $S$-velocities $\alpha_h$,
    $\beta_h$ at source depth $h$ \citep{AkiRichards1980}. The expressions are:

@@

    We thus need a metric against which to evaluate what the best and acceptable estimates are. The
    misfit function used by \citet{sita_potential_2022} is defined as the angle between the vector of
    observed amplitudes and that produced from a forward simulation of synthetic parameters.
    In this paper, we shall use a misfit $\mu$ of cosine similarity between the two vectors which has a
    1-to-1 correspondence with angle:

    \begin{align} \label{eq:eq3}
      \mu(\bm{p}) = \frac{\bm{A}_o^\top\bm{A}_s}{\|\bm{A}_o\|\|\bm{A}_s\|}
    \end{align}

    where $\bm{A}_o = (A_o^P, A_o^{SV}, A_o^{SH})^\top$ are observed amplitudes, $\bm{A}_s =
    (A_s^P, A_s^{SV}, A_s^{SH})^\top$ are synthetic amplitudes, and
    $\bm{p} = (\psi, \delta, \lambda)^\top$ are the underlying parameters.

    \subsection{Justification}
    We are interested in situations where absolute amplitudes are either unknown or unnecessary, for
    instance if using data collected from NASA's Insight mission \citep{sita_potential_2022}.
    In this case, we cannot know the absolute velocities because they depend on an understanding of
    Mars' interior that is good enough to build a reliable velocity model. Alternative misfit
    functions can account for the relative amplitudes, such as in equations \ref{eq:eq4} and
    \ref{eq:eq5}:

    \begin{align} \label{eq:eq4}
      \mu(\bm{p}) = \sqrt{\left( \Delta \frac{A^{P}}{A^{SV}}\right)^2 + \left( \Delta \frac{A^{SV}}{A^{SH}}\right)^2}
    \end{align}

    \begin{align} \label{eq:eq5}
      \mu(\bm{p}) = \sqrt{\log^2 \left( \Delta \frac{A^P}{A^{SV}} \right) + \log^2 \left( \Delta \frac{A^{SV}}{A^{SH}} \right)}
    \end{align}

    First, we note that using cosine similarity is more numerically stable in cases where one of the
    observed or synthetic amplitudes is close to zero since the norm of the associated vector in
    equation \ref{eq:eq3} may still be large enough to make division feasible.

    Secondly, we have chosen cosine similarity over angle because it involves one less operation,
    which is more computationally efficient and easier to differentiate for directed search
    algorithms (see Section \ref{sec:algorithms}). The Gaussian function in equation \ref{eq:eq6} is
    similarly easy to differentiate, but it does not account for relative amplitudes so it would
    exclude otherwise acceptable solutions.

    \begin{align} \label{eq:eq6}
      &\mu(\bm{p}) \sim |\Sigma|^{-1/2} \exp \left( - \frac{1}{2} (\bm{A} - \bm{A}_o)^\top \bm{\Sigma}^{-1} (\bm{A} - \bm{A}_o) \right)
    \end{align}

    \subsection{Tolerance}
    Given vectors $\bm{A}_s$ and $\bm{A}_o$, we would like to know the largest cosine
    similarity for which $\bm{A}_s$ is considered an acceptable fit with respect to the recorded amplitudes
    while accounting for the asymmetry of error levels for different amplitudes. Visualized in 3D
    space, we have a confidence ellipsoid centered at $\bm{A}_o$ (equation \ref{eq:eq7}) such that
    acceptable fits are vectors which fall inside the elliptical cone tangent to it with vertex at the
    origin.
    \begin{align} \label{eq:eq7}
      E(\bm{p}) = (\bm{A} - \bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = 1
    \end{align}

    \noindent where $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ is arbitrary
    and $\hat{\bm{\Sigma}} = \text{diag}(\sigma_P^2, \sigma_{SV}^2, \sigma_{SH}^2)/\chi^2_{2,\alpha}$ for some $\alpha \in (0,1)$. The elliptical
    cone intersects with the ellipsoid at every boundary point for which
    \begin{align} \label{eq:eq8}
      (\nabla_{\bm{A}} E)^\top \bm{A} = 2(\bm{A} - \bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} \bm{A} = 0
    \end{align}

    Combining equations \ref{eq:eq7} and \ref{eq:eq8} gives the plane containing every point of 
    tangency:
    \begin{align} \label{eq:eq9}
      \frac{1}{2} (\nabla_{\bm{A}} E)^\top \bm{A} - f(\bm{A}) = 
              \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = -1
    \end{align}

    To account for asymmetry, we recognize that the orientation of $\bm{A}_s$ relative to $\bm{A}_o$
    has an effect on the tolerance because the associated "thickness" of the ellipsoid changes.
    This information is captured by the plane containing vectors $\bm{A}_o$ and $\bm{A}_s$:
    \begin{align} \label{eq:eq10}
      (\bm{A}_s \times \bm{A}_o)^\top \bm{A} = 0
    \end{align}

    To find the constrained points of tangency, we can intersect the original ellipsoid with planes
    \ref{eq:eq9} and \ref{eq:eq10}. The line of intersection is such that
    \begin{align} \label{eq:eq11}
      \bm{A}(t) = (1 - \nicefrac{1}{c_1}) \bm{A}_o + t\bm{D}
    \end{align}

    \noindent where $c_1$ (equation \ref{eq:eq12}) is a location parameter projecting $\bm{A}_o$ onto plane \ref{eq:eq9}
    and $\bm{D}$ (equation \ref{eq:eq13}) is the direction perpendicular to planes \ref{eq:eq9} and \ref{eq:eq10}.

    \begin{align} \label{eq:eq12}
      \nonumber \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} ((1 - \nicefrac{1}{c_1})\bm{A}_o - \bm{A}_o)  = -1\\
      \Rightarrow c_1 = \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{A}_o = \|\bm{A}_o^\top, \bm{A}_o\|_{\hat{\bm{\Sigma}}^{-1}}^2
    \end{align}
    \begin{align} \label{eq:eq13}
      \nonumber\bm{D} &= (\bm{A}_o^\top \hat{\bm{\Sigma}}^{-1})^\top \times (\bm{A}_s \times \bm{A}_o)\\
        &= c_1\bm{A}_s - c_2\bm{A}_o
    \end{align}

    \noindent where $c_2 = \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{A}_s = \langle\bm{A}_o^\top, \bm{A}_s\rangle_{\hat{\bm{\Sigma}}^{-1}}$.

    Combining equations \ref{eq:eq7} and \ref{eq:eq11} produces a quadratic parameterized by the scalar $t$,
    from which we can obtain a quadratic equation for the points of tangency:

    \begin{align} \label{eq:eq14}
      \nonumber (t\bm{D} - \nicefrac{1}{c_1}\bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} (t\bm{D} - \nicefrac{1}{c_1}\bm{A}_o)  = 1\\
      \nonumber ({\bm{D}^\top \hat{\bm{\Sigma}}^{-1} \bm{D}})t^2 - \nicefrac{2}{c_1}(\bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{D})t + (\nicefrac{1}{c_1} - 1) = 0\\
      % \Rightarrow t = \nicefrac{c_3}{c_1c_2} \pm \nicefrac{1}{c_1c_2} \sqrt{c_3^2 + c_1c_2(c_1 - 1)}
    \end{align}

    Denote $c_3 = \bm{A}_s^\top \hat{\bm{\Sigma}}^{-1} \bm{A}_s = \|\bm{A}_s^\top, \bm{A}_s\|_{\hat{\bm{\Sigma}}^{-1}}^2$. Then by
    substitution we have
    ${\bm{D}^\top \hat{\bm{\Sigma}}^{-1} \bm{D}} = c_1(c_1c_3 - c_2^2)$ and ${\bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{D}} = 0$.
    Solving then yields the necessary points (denoted $\bm{A}_b$) as
    \begin{align} \label{eq:eq15}
      \bm{A}_b = \left(1 - \frac{1}{c_1}\right)\bm{A}_o \pm
        \left(\frac{1}{c_1} \sqrt{\frac{c_1 - 1}{c_1c_3 - c_2^2}}\right)(c_1\bm{A}_s - c_2\bm{A}_o)
    \end{align}

    Equation \ref{eq:eq15} shows that for the points to be well-defined, the following conditions
    must hold relative to the inner-product space defined by $\hat{\bm{\Sigma}}^{-1}$:

    \begin{itemize}
      \item Strict Cauchy-Schwarz inequality: $c_1c_3 > c_2^2$. This also implies that
        $\hat{\bm{\Sigma}} \succ 0$ and $\bm{A}_s \nparallel \bm{A}_o$. Practically, it means that all
        uncertainties are necessary and inversion methods that depend on forward simulation should
        treat parallelism as an edge case.
      \item Confidence ellipsoid cannot contain origin: $c_1 \geq 1$. Practically, we should expect
      errors to be a small fraction of observed amplitudes.
    \end{itemize}

    Thus, given 2 nonparallel vectors $\bm{A}_o$ and $\bm{A}_s$ and positive definite covariance
    matrix $\hat{\bm{\Sigma}}$, we accept the simulation if $\hat{\bm{A}}_o^\top \hat{\bm{A}_s} \geq \text{tol} 
    = \min_b \hat{\bm{A}}_o^\top \hat{\bm{A}_b}$. The condition, more formally, is
    \begin{align} \label{eq:eq16}
      \max_b \hat{\bm{A}}_o^\top (\hat{\bm{A}}_b - \hat{\bm{A}}_s) \leq 0
    \end{align}

    \noindent where $\hat{\bm{A}}_o$, $\hat{\bm{A}}_s$ and $\hat{\bm{A}}_b$ are unit vectors.

@@

    The vector $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ can be treated as a continuously differentiable nonlinear function of
    parameters $\bm{p} = (\psi, \delta, \lambda)^\top$, defined by equation \ref{eq:eq1}. We can thus apply
    directed search methods to minimize the following cost function:
    \begin{align} \label{eq:eq17}
      \mu(\bm{p}) = \hat{\bm{A}}_o^\top \hat{\bm{A}}
    \end{align}

    TODO: draw computational graph, derive gradients
